Synthesizing Quotient Lenses

### 2017-12-15 papers/QuotientLenses/main.tex

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
  array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\lstset{
  language=Caml,
  basicstyle=\upshape\sffamily,
  keywordstyle=\upshape\sffamily\color{dkblue},
  keepspaces=true,
  framexleftmargin=1ex,
  framexrightmargin=1ex,
  showstringspaces=false,
  commentstyle=\itshape\rmfamily,
  columns=flexible,
  % BCP: I find this distracting:
  % stringstyle=\sffamily\color{dkred},
  mathescape
}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

@@

\newcommand{\OpticianRuntime}{\textbf{OO}}
\newcommand{\SystemOnOptician}{\textbf{SO}}
\newcommand{\SystemOnBenchmarks}{\textbf{SS}}

@@

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{Boomerang} is the only implementation
currently offering quotient lenses.  One reason for this is that there is
some annoying redundancy in the quotient lens notation: The programmer often
has to separately specify (1) the data formats on each side of the lens, (2) the
equivalence relations over them, and (3) a lens between the canonical
representatives of the equivalence classes, even though, typically, these three
share a great deal of structure.  Moreover, writing the lens between the
equivalence classes can be challenging due to the rather fiddly typechecking,
mainly due to ``non-ambiguity'' conditions in the lens well-formedness rules.

@@

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QRE
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations that are
difficult to express, or that cannot be expressed, using the other
combinators.

However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable in general, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can use any lens
to canonize a regular language described by $R$ to a regular language described
by $S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE
$c$ canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$. QREs therefore are
not as expressive as Boomerang canonizers. On the other hand, QREs enable us to
express many equivalences which occur in real world data as we shall demonstrate
in the evaluation section (Section~\ref{impl}).

@@

\subsection{Well Formed QREs and ambiguity}
Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity} which comes up when defining QREs. If $c$ and  $c'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(c) \cdot W(c')$, $W(c) \sep W(c')$,
$W(c)^*$, $K(c) \cdot K(c')$, $K(c) \sep K(c')$, $K(c)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}, and is necessary, firstly because it ensures that the
canonizing function of a QRE is well-defined.

@@

\noindent The behaviour of \cd{ambiguous} is not-well defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a", "aa", "aaa"} depending on how
\cd{"aaa"} is parsed. We also require that the regular combinators applied to
the kernels are unambiguous since the underlying lenses will end up operating
on the kernels, and these lenses impose the same restrictions for similar
reasons.

To this end, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$.

@@

\begin{figure}[p!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\mathit{id}(R)$ \textbf{WF}}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\UnaryInfC{$\project{R}{s}$ \textbf{WF}}
\DisplayProof
\hskip 1.5em 
\end{center}

\begin{prooftree}
\AxiomC{$c_i, c$ are well formed}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(c_{\sigma(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\sigma(n)})\\ \cap W(c_{\theta(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(c_1) \cdot^! K(c) \cdot^! \ldots \cdot^! K(c) \cdot^! K(c_n)$}
\TrinaryInfC{$\perm{c_1, \ldots, c_n}{c}$ \textbf{WF}}
\end{prooftree}

\begin{center}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$\mathcal{L}(W(c)) \cap \mathcal{L}(W(c')) = \varnothing$}
\AxiomC{$f : \mathcal{L}(W(c)) \longrightarrow \mathcal{L}(W(c'))$}
\QuaternaryInfC{$\squash{c}{c'}{f}$ \textbf{WF}}
\DisplayProof
\end{center}
\begin{center}

\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\QuaternaryInfC{$\normalize{R}{R'}{f}$ \textbf{WF}}
\end{prooftree}

\end{center}
\begin{center}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$K(c) = W(c')$}
\TrinaryInfC{$c' \; ; \; c$ \textbf{WF}}
\DisplayProof
\hskip 1.5em
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$W(c)^{*!}$}
\AxiomC{$K(c)^{*!}$}
\TrinaryInfC{$c^*$ \textbf{WF}}
\DisplayProof
\end{center}
\begin{center}

\begin{center}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\QuaternaryInfC{$c \cdot c'$ \textbf{WF}}
\DisplayProof
\end{center}

\begin{center}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$W(c) \cap W(c') = \varnothing$}
\TrinaryInfC{$c \cdot c$ \textbf{WF}}
\DisplayProof
\hskip 1.5em
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$W(c)^{*!}$}
\AxiomC{$K(c)^{*!}$}
\TrinaryInfC{$c^*$ \textbf{WF}}
\DisplayProof
\end{center}
\end{center}
  \caption{Well-formed QREs 
}
  \label{fig:qrerules}
\end{figure}

Figure~\ref{fig:qrerules} gives the inferrence rules for deriving well-formed
QREs. As expected, the unambiguity conditions come into play when defining QREs
using the regular combinators. For example, the inferrence rule for
concatenation is as follows:
\begin{prooftree}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$c'$ \textbf{WF}}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\QuaternaryInfC{$c \cdot c'$ \textbf{WF}}
\end{prooftree}

@@

\begin{prooftree}
\AxiomC{$c_i, c$ are well formed}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(c_{\sigma(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\sigma(n)})\\ \cap W(c_{\theta(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(c_1) \cdot^! K(c) \cdot^! \ldots \cdot^! K(c) \cdot^! K(c_n)$}
\TrinaryInfC{$\perm{c_1, \ldots, c_n}{c}$ \textbf{WF}}
\end{prooftree}

@@

The rest of the inferrence rules for deriving QREs are given in
Figure~\ref{fig:qrerules}.
\subsection{Semantics of QREs}

As we mentioned before, each QRE $c$ enables us to express:
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
  \end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole language
$W(c)$ of a QRE. The \textit{squash} and \textit{permutation} combinators
give the two interesting definitions . If $c = \squash{c}{c'}{f}$, then
the whole language of $c$ is $W(c) \sep W(c')$. This is because the
\textit{squash} combinator takes the whole language $W(c)$ of $c$, the whole
language $W(c')$ of $c'$ and a function $f : \mathcal{L}(W(c)) \longrightarrow
\mathcal{L}(W(c'))$, maps $\mathcal{L}(W(c))$ into $\mathcal{L}(W(c'))$ using
$f$ and then canonizes $W(c')$ into $K(c')$ using $\canonizer(c')$.

@@

      c & W(c) & K(c) \\ \hline
      R & R & R \\
      \project{R}{s} & R & s \\
      \squash{c}{c'}{f} & W(c) \sep W(c') & K(c') \\
      \normalize{R}{R'}{f} & R & R' \\
      c_2 \; ; \; c_1 & W(c_1) & K(c_2) \\
      c_1 \cdot c_2 & W(c_1) \cdot W(c_2) & K(c_1) \cdot K(c_2) \\
      c_1 \sep c_2 & W(c_1) \sep W(c_2) & K(c_1) \sep K(c_2) \\
      c^* & W(c)^* & K(c)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{c_1, \ldots, c_n}{c} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})\\
K( \perm{c_1, \ldots, c_n}{c} ) = & K(c_1) \cdot K(c) \cdot \ldots \cdot K(c)
\cdot K(c_n)
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions\bcp{try putting the 
      last two clauses into the same format as the
      others (using $\mathit{qq} = ...$ and maybe reducing spacing around
      cdots, etc.)}}
  \label{fig:wk}
\end{figure}

@@

 \begin{figure}[ht]
  \centering
  \begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\UnaryInfC{$\kw{lift}(\ell): id(R) \Leftrightarrow id(S)$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &=  \llbracket \ell \rrbracket\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1}
  \end{align*}

\begin{prooftree}
\AxiomC{$q' : c'  \Leftrightarrow c''$}
\AxiomC{$c$ \textbf{WF}}
\AxiomC{$K(c) = W(c')$}
\TrinaryInfC{$\kw{lquot}(c, q'): c' \circ c \Leftrightarrow c''$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get  &= \llbracket q'
  \rrbracket.get \circ \canonizer(c)\\
  \llbracket q \rrbracket.put &= \llbracket q' \rrbracket.put
  \end{align*}

  \begin{prooftree}
  \AxiomC{$q' : c \Leftrightarrow c'$}
  \AxiomC{$c''$ \textbf{WF}}

\AxiomC{$K(c'') = W(c')$}
\TrinaryInfC{$\kw{rquot}(q', c):c \Leftrightarrow c'' \circ c'$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &= \llbracket q'
  \rrbracket.get\\
  \llbracket q \rrbracket.put &= \llbracket q'
  \rrbracket.put \circ \canonizer(c'')
  \end{align*}

@@

    \begin{prooftree}
\AxiomC{$q : c \Leftrightarrow c$}
\AxiomC{$\substack{W(c)^{*!} \text{ and } W(c')^{*!} \\ K(c)^{*!} \text{ and }
K(c')^{*!}}$}
\BinaryInfC{${q}^* : c^* \Leftrightarrow {c'}^*$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &= (\llbracket q \rrbracket.get)^* \\
  \llbracket q \rrbracket.put &= (\llbracket q \rrbracket.put)^*
  \end{align*}

    \begin{prooftree}
\AxiomC{$\substack{q_1 : c_1 \Leftrightarrow d_1 \\ q_2 : c_2 \Leftrightarrow
d_2}$}
\AxiomC{${\substack{W(c_1) \cdot^! W(c_2)\\ K(c_1) \cdot^! K(c_2)}}$}
\AxiomC{${\substack{W(d_1) \cdot^! W(d_2)\\ K(d_1) \cdot^! K(d_2)}}$}
\TrinaryInfC{$q_1 \cdot q_2: c_1 \cdot c_2
\Leftrightarrow d_1 \cdot d_2$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &= \llbracket q_1 \rrbracket.get \cdot \llbracket
  q_2 \rrbracket.get \\
  \llbracket q \rrbracket.put &= \llbracket q_1 \rrbracket.put \cdot \llbracket
  q_2 \rrbracket.put
  \end{align*}

      \begin{prooftree}
\AxiomC{$\substack{q_1 : c_1 \Leftrightarrow d_1 \\ q_2 : c_2 \Leftrightarrow
d_2}$}
\AxiomC{$\substack{\mathcal{L}(W(c_1)) \cap \mathcal{L}(W(c_2)) = \varnothing
\\ \mathcal{L}(W(d_1)) \cap \mathcal{L}(W(d_2)) = \varnothing}$}
\BinaryInfC{$q_1 \sep q_2: (c_1 \sep c_2)
\Leftrightarrow (d_1 \sep d_2)$}
\end{prooftree}
  $$
  \llbracket q_1 \sep q_2 \rrbracket.get(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_1))\\
  \llbracket q_2 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_2))\\
  \end{cases}$$
  $$\llbracket q_1 \sep q_2 \rrbracket.put(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_1))\\
  \llbracket q_2 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_2))\\
  \end{cases}
  $$
  \caption{Denotation and Typing Rules for QRE Lenses\bcp{I think we can
      display this more compactly if we make the inference rules narrower by
    stacking the premises vertically and then put each inference rule {\em
      beside} its semantics definition.}}
  \label{fig:qlenssemantics}
\end{figure}

>>>>

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
  array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\usepackage{proof}

\lstset{
  language=Caml,
  basicstyle=\upshape\sffamily,
  keywordstyle=\upshape\sffamily\color{dkblue},
  keepspaces=true,
  framexleftmargin=1ex,
  framexrightmargin=1ex,
  showstringspaces=false,
  commentstyle=\itshape\rmfamily,
  columns=flexible,
  % BCP: I find this distracting:
  % stringstyle=\sffamily\color{dkred},
  mathescape
}

@@

% For Inference Rules
\newcommand{\Rule}[2]{\infer{#2}{#1}}
\newcommand{\RuleSide}[3]{\infer[#3]{#2}{#1}}
%\newcommand{\Axiom}[1]{\Rule{}{#1}}

\newcommand{\wf}[1]{\ensuremath{#1\;\mathsf{wf}}}

@@

\newcommand{\OpticianRuntime}{\textbf{OO}}
\newcommand{\SystemOnOptician}{\textbf{SO}}
\newcommand{\SystemOnBenchmarks}{\textbf{SS}}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{boomerang} is the only implementation
currently offering quotient lenses.  One reason for this is that there is
some annoying redundancy in the quotient lens notation: The programmer often
has to separately specify (1) the data formats on each side of the lens, (2) the
equivalence relations over them, and (3) a lens between the canonical
representatives of the equivalence classes, even though, typically, these three
share a great deal of structure.  Moreover, writing the lens between the
equivalence classes can be challenging due to the rather fiddly typechecking,
mainly due to ``non-ambiguity'' conditions in the lens well-formedness rules.

@@

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QRE
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations that are either 
difficult to define, or that cannot be defined using the other combinators. 
\saz{Add something about being ``complete'' here?}\sam{I'm not sure how to do
this here. The precise statement about how normalize is strong enough to
express all QREs requires that we introduce W(c), K(c) and Canonizer(c) which
we haven't introduced at this point in the paper}

\saz{Say more about how normalize
wouldn't be used in ``ordinary'' QRE programming?}\sam{I'm not sure what you
mean here exactly\ldots} However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable in general, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can use any lens
to canonize a regular language described by $R$ to a regular language described
by $S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE
$c$ canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$. QREs therefore are
not as expressive as Boomerang canonizers. On the other hand, QREs enable us to
express many equivalences that occur in real world data as we shall demonstrate
in the evaluation section (Section~\ref{impl}).

@@

\subsection{Well Formed QREs and ambiguity}
Before giving the formal semantics for QREs, we address the issue of {\em
ambiguity}, which arises when defining QREs. If $c$ and  $c'$ are QREs, then
when applying the regular combinators to QREs, we require that if a string $s$
matches any of the regular expressions $W(c) \cdot W(c')$, $W(c) \sep W(c')$,
$W(c)^*$, $K(c) \cdot K(c')$, $K(c) \sep K(c')$, $K(c)^*$, then $s$ matches
that regular expression in only one way. This unambiguity condition is called
{\em strong unambiguity}~\cite{Sippu1988}, and is necessary, firstly because it
ensures that the canonizing function of a QRE is well-defined.

@@

\noindent The behaviour of \cd{ambiguous} is not-well defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"} depending on how
\cd{"aaa"} is parsed.

We also require that the regular combinators applied to
the kernels are unambiguous since the underlying lenses will end up operating
on the kernels, and these lenses impose the same restrictions for similar
reasons.
%
To this end, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$.

@@

\begin{figure}[p!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\wf{\mathit{id}(R)}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\UnaryInfC{$\wf{\project{R}{s}}$}
\DisplayProof
\hskip 1.5em 
\end{center}

\begin{prooftree}
\AxiomC{$\wf{c_i, c}$}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(c_{\sigma(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\sigma(n)})\\ \cap W(c_{\theta(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(c_1) \cdot^! K(c) \cdot^! \ldots \cdot^! K(c) \cdot^! K(c_n)$}
\TrinaryInfC{$\wf{\perm{c_1, \ldots, c_n}{c}}$}
\end{prooftree}

\begin{center}
\AxiomC{$\wf{c, c'}$}
\AxiomC{$\mathcal{L}(W(c)) \cap \mathcal{L}(W(c')) = \varnothing$}
\AxiomC{$f : \mathcal{L}(W(c)) \longrightarrow \mathcal{L}(W(c'))$}
\TrinaryInfC{$\wf{\squash{c}{c'}{f}}$}
\DisplayProof
\end{center}
\begin{center}

\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}

\end{center}
\begin{center}
\AxiomC{$\wf{c, c'}$}
\AxiomC{$K(c) = W(c')$}
\BinaryInfC{$\wf{c' \; ; \; c}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\wf{c}$}
\AxiomC{$W(c)^{*!}$}
\AxiomC{$K(c)^{*!}$}
\TrinaryInfC{$\wf{c^*}$}
\DisplayProof
\end{center}
\begin{center}

\begin{center}
\AxiomC{$\wf{c, c'}$}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\TrinaryInfC{$\wf{c \cdot c'}$}
\DisplayProof
\end{center}

\begin{center}
\AxiomC{$\wf{c, c'}$}
\AxiomC{$W(c) \cap W(c') = \varnothing$}
\BinaryInfC{$\wf{c \cdot c}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\wf{c}$}
\AxiomC{$W(c)^{*!}$}
\AxiomC{$K(c)^{*!}$}
\TrinaryInfC{$\wf{c^*}$}
\DisplayProof
\end{center}
\end{center}
  \caption{Well-formed QREs 
}
  \label{fig:qrerules}
\end{figure}

Figure~\ref{fig:qrerules} gives the inferrence rules for deriving well-formed
QREs. As expected, the unambiguity conditions come into play when defining QREs
using the regular combinators. For example, the inferrence rule for
concatenation is as follows:
\begin{prooftree}
\AxiomC{$\wf{c, c'}$}
\AxiomC{$W(c) \cdot^! W(c')$}
\AxiomC{$K(c) \cdot^! K(c')$}
\TrinaryInfC{$\wf{c \cdot c'}$}
\end{prooftree}

@@

\begin{prooftree}
\AxiomC{$\wf{c_i, c}$}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(c_{\sigma(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\sigma(n)})\\ \cap W(c_{\theta(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(c_1) \cdot^! K(c) \cdot^! \ldots \cdot^! K(c) \cdot^! K(c_n)$}
\TrinaryInfC{$\wf{\perm{c_1, \ldots, c_n}{c}}$}
\end{prooftree}

@@

The rest of the inferrence rules for deriving QREs are given in
Figure~\ref{fig:qrerules}.
\subsection{Semantics of QREs}
As we mentioned before, each QRE $c$ enables us to express the whole and kernel
languages $W(c)$ and $K(c)$, an equivalence relation $\eqrel{c}$ on
$\mathcal{L}(W(c))$ that has $\mathcal{L}(K(c))$ as a complete set of
representatives and a canonizing function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ whose fibres are the same as $\eqrel{c}$.

Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(c)$ and $K(c)$ of a QRE. The \textit{squash} and
\textit{permutation} combinators give the two interesting definitions. If $c =
\squash{c}{c'}{f}$, then the whole language of $c$ is $W(c) \sep W(c')$. This
is because the \textit{squash} combinator takes the whole language $W(c)$ of
$c$, the whole language $W(c')$ of $c'$ and a function $f : \mathcal{L}(W(c))
\longrightarrow \mathcal{L}(W(c'))$, maps $\mathcal{L}(W(c))$ into
$\mathcal{L}(W(c'))$ using $f$ and then canonizes $W(c')$ into $K(c')$ using
$\canonizer(c')$.

@@

      c & W(c) & K(c) \\ \hline
      R & R & R \\
      \project{R}{s} & R & s \\
      \squash{c}{c'}{f} & W(c) \sep W(c') & K(c') \\
      \normalize{R}{R'}{f} & R & R' \\
      c_2 \; ; \; c_1 & W(c_1) & K(c_2) \\
      c_1 \cdot c_2 & W(c_1) \cdot W(c_2) & K(c_1) \cdot K(c_2) \\
      c_1 \sep c_2 & W(c_1) \sep W(c_2) & K(c_1) \sep K(c_2) \\
      c^* & W(c)^* & K(c)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{c_1, \ldots, c_n}{c} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})\\
K( \perm{c_1, \ldots, c_n}{c} ) = & K(c_1) \cdot K(c) \cdot \ldots \cdot K(c)
\cdot K(c_n)
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions}
  \label{fig:wk}
\end{figure}

@@

 \begin{figure}[ht]
  \centering
\[
\begin{array}[b]{l@{\qquad}l}
\Rule{\ell : R \Leftrightarrow S}{\kw{lift}(\ell): id(R) \Leftrightarrow id(S)}
&
  \begin{array}[b]{rcl}
  \llbracket \kw{lift}(\ell) \rrbracket.get &=&  \llbracket \ell \rrbracket\\
  \llbracket \kw{lift}(\ell) \rrbracket.put &=& \llbracket \ell \rrbracket^{-1}
  \end{array}
  \end{array}
  \]
  \[
\begin{array}[b]{c@{\qquad}l}
  \Rule{q : c'  \Leftrightarrow c'' &
 \wf{c} &
 K(c) = W(c')}
{\kw{lquot}(c, q): c' \circ c \Leftrightarrow c''} &
  \begin{array}[b]{rcl}
  \llbracket \kw{lquot}(c, q) \rrbracket.get  &=& \llbracket q
  \rrbracket.get \circ \canonizer(c)\\
  \llbracket \kw{lquot}(c, q) \rrbracket.put &=& \llbracket q \rrbracket.put
  \end{array}
\end{array}
\]

\[
\begin{array}[b]{c@{\qquad}c}
\Rule{q : c'  \Leftrightarrow c'' &
 \wf{c} &
 K(c) = W(c')}
{\kw{lquot}(c, q): c' \circ c \Leftrightarrow c''} &
  \begin{array}[b]{rcl}
  \llbracket \kw{lquot}(c, q) \rrbracket.get  &=& \llbracket q
  \rrbracket.get \circ \canonizer(c)\\
  \llbracket \kw{lquot}(c, q) \rrbracket.put &=& \llbracket q \rrbracket.put
  \end{array}
\end{array}
\]

\[
\begin{array}[b]{c@{\qquad}c}
\Rule{q : c \Leftrightarrow c'' &
 \wf{c''} &
 W(c'') = K(c')}
{\kw{rquot}(q, c):c \Leftrightarrow c'' \circ c'} &
  \begin{array}[b]{rcl}
  \llbracket \kw{rquot}(q, c) \rrbracket.get  &=& \llbracket q
  \rrbracket.get\\
  \llbracket \kw{rquot}(q, c) \rrbracket.put &=& \llbracket q
  \rrbracket.put \circ \canonizer(c'')
  \end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\Rule{q : c \Leftrightarrow c' &
 W(c)^{*!},W(c')^{*!} & K(c)^{*!}, K(c')^{*!}}
{{q}^* : c^* \Leftrightarrow {c'}^*} &
  \begin{array}[b]{rcl}
  \llbracket {q}^* \rrbracket.get  &=& (\llbracket q \rrbracket.get)^*\\
  \llbracket {q}^* \rrbracket.put &=& (\llbracket q \rrbracket.get)^*
  \end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\Rule{\substack{q_1 : c_1 \Leftrightarrow d_1 \\ q_2 : c_2 \Leftrightarrow
d_2} & {\substack{W(c_1) \cdot^! W(c_2)\\ K(c_1) \cdot^! K(c_2)}} 
& {\substack{W(d_1) \cdot^! W(d_2)\\ K(d_1) \cdot^! K(d_2)}}}
{q_1 \cdot q_2: c_1 \cdot c_2 \Leftrightarrow d_1 \cdot d_2} &
  \begin{array}[b]{rcl}
  \llbracket q_1 \cdot q_2 \rrbracket.get  &=& \llbracket q_1 \rrbracket.get
  \cdot \llbracket q_2 \rrbracket.get\\
  \llbracket q_1 \cdot q_2\rrbracket.put &=& \llbracket q_1 \rrbracket.put
  \cdot \llbracket q_2 \rrbracket.put
  \end{array}
\end{array}
\]

\[
\begin{array}[b]{l@{\qquad}c}
\Rule{\substack{q_1 : c_1 \Leftrightarrow d_1 \\ q_2 : c_2 \Leftrightarrow
d_2} & \substack{\mathcal{L}(W(c_1)) \cap \mathcal{L}(W(c_2)) = \varnothing
\\ \mathcal{L}(W(d_1)) \cap \mathcal{L}(W(d_2)) = \varnothing }}
{q_1 \sep q_2: (c_1 \sep c_2)\Leftrightarrow (d_1 \sep d_2)} &
  \begin{array}[b]{rcl}
  \llbracket q_1 \sep q_2 \rrbracket.get(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_1))\\
  \llbracket q_2 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_2))\\
  \end{cases}\\
  \llbracket q_1 \sep q_2 \rrbracket.put(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_1))\\
  \llbracket q_2 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_2))\\
  \end{cases}
  \end{array}
\end{array}
\]

@@

  \caption{Denotation and Typing Rules for QRE Lenses}
  \label{fig:qlenssemantics}
\end{figure}
